---
title: "In-class Exercise 6: Horizon Plot"
author: "Si Min Kock"
date: "February 24, 2024"
date-modified: "last-modified"
toc: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
editor: visual
---

## 1. Installing R Packages

```{r}
pacman::p_load(ggHoriPlot,ggthemes,tidyverse)
```
## 2. Importing Data

```{r}
averp <- read_csv("data/AVERP.csv") %>%
  mutate(`Date`=dmy(`Date`))
```

::: {.callout-note}
Date/Time data is always tricky!

Always remember to check whether date data is in correct format. Prior to mutating the date above, the date was recognised as a character data type.

`dmy()` of [`lubridate`](https://lubridate.tidyverse.org/) package will parse the date field into the appropriate data type in R.
:::

## 3. Plotting the horizon plot

```{r}
#| fig-width: 12
#| fig-height: 10
averp %>% 
  filter(Date >= "2018-01-01") %>%
  ggplot() + # don't need to indicate data here as it is in the first line of code
  geom_horizon(aes(x = Date, y=Values), 
               origin = "midpoint", 
               horizonscale = 6)+
  facet_grid(`Consumer Items`~.) +
    theme_few() + #theme to use less features
  scale_fill_hcl(palette = 'RdBu') + # use diverging color scheme only when you have positive and negative data to get good separation however if looking at just positive data i.e. just need to observe intensity of data, just use single color
  theme(panel.spacing.y=unit(0, "lines"), strip.text.y = element_text(
    size = 5, angle = 0, hjust = 0),
    legend.position = 'none',
    axis.text.y = element_blank(),
    axis.text.x = element_text(size=7),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.y = element_blank(),
    panel.border = element_blank()
    ) +
    scale_x_date(expand=c(0,0), date_breaks = "3 month", date_labels = "%b%y") +
  ggtitle('Average Retail Prices of Selected Consumer Items (Jan 2018 to Dec 2022)')
```

::: {.callout-note}
One can configure the figure width and height by including lines of code "fig-width:12" and "fig-height:10" above before the main code.
:::