---
title: "Hands-on Exercise 4B"
date: January 28, 2024
date-modified: "last-modified"
toc: true
execute:
  eval: true
  echo: true
  warning: false
---

# 4 Visual Statistical Analysis

## 4.1 Learning Outcome

In this hands-on exercise, I will be using:

-   [ggstatsplot](https://indrajeetpatil.github.io/ggstatsplot/) package to create visual graphics with rich statistical information,

-   [performance](https://easystats.github.io/performance/) package to visualise model diagnostics, and

-   [parameters](https://easystats.github.io/parameters/) package to visualise model parameters

## 4.2 Visual Statistical Analysis with ggstatsplot

-   [**ggstatsplot**](https://indrajeetpatil.github.io/ggstatsplot/) is an extension of [**ggplot2**](https://ggplot2.tidyverse.org/) package for creating graphics with details from statistical tests included in the information-rich plots themselves.

    -   To provide alternative statistical inference methods by default.

    -   To follow best practices for statistical reporting. For all statistical tests reported in the plots, the default template abides by the [APA](https://my.ilstu.edu/~jhkahn/apastats.html) gold standard for statistical reporting.

## 4.3 Getting Started

## 4.3.1 Installing and launching R packages

In this exercise, **ggstatsplot** and **tidyverse** will be used.

```{r}
pacman::p_load(ggstatsplot,tidyverse)
```

### 4.3.2 Importing data

```{r}
exam <- read_csv("data/Exam_data.csv")
exam
```

### 4.3.3 One-sample test: *gghistostats()* method

In the code chunk below, [*gghistostats()*](https://indrajeetpatil.github.io/ggstatsplot/reference/gghistostats.html) is used to to build an visual of one-sample test on English scores.

```{r}
#| eval: false
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores",
  title="Distribution of English scores"
)
```

```{r,echo=FALSE}
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores",
  title="Distribution of English scores"
)
```

Default information: - statistical details - Bayes Factor - sample sizes - distribution summary

### 4.3.4 Unpacking the Bayes Factor

-   A Bayes factor is the ratio of the likelihood of one particular hypothesis to the likelihood of another. It can be interpreted as a measure of the strength of evidence in favor of one theory among two competing theories.

-   That’s because the Bayes factor gives us a way to evaluate the data in favor of a null hypothesis, and to use external information to do so. It tells us what the weight of the evidence is in favor of a given hypothesis.

-   When we are comparing two hypotheses, H1 (the alternate hypothesis) and H0 (the null hypothesis), the Bayes Factor is often written as B10.

<!-- -->

-   The [**Schwarz criterion**](https://www.statisticshowto.com/bayesian-information-criterion/) is one of the easiest ways to calculate rough approximation of the Bayes Factor.

### 4.3.5 How to interpret Bayes Factor

A **Bayes Factor** can be any positive number. 

### 4.3.6 Two-sample mean test: *ggbetweenstats()*

In the code chunk below, [*ggbetweenstats()*](https://indrajeetpatil.github.io/ggstatsplot/reference/ggbetweenstats.html) is used to build a visual for two-sample mean test of Maths scores by gender.

```{r}
#| eval: false
ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE,
  title="Distribution of Maths Scores by Gender"
)
```

```{r,echo=FALSE}
ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE,
  title="Distribution of Maths Scores by Gender"
)
```
Default information: - statistical details - Bayes Factor - sample sizes - distribution summary

