---
title: "Hands-on Exercise 1"
date: January 6, 2024
date-modified: "last-modified"
toc: true
execute:
  eval: true
  echo: true
  warning: false
---

# Getting Started

## Loading R package

```{r}
pacman::p_load(tidyverse)

```

# Importing Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Wrangling Data

### Geom bar

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()+
  ggtitle("Distribution of cohort of primary 3 students from a local school by race")
```

### Geom dotplot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,
               dotsize = 0.5)+
  scale_y_continuous(NULL,
                     breaks=NULL)+
  ggtitle("Distribution of students by MATHS score")
```

### Geom histogram

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light yellow")+
  ggtitle("Distribution of students by MATHS score")

ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")+
  ggtitle("Distribution of students by MATHS score and by gender")
```

### Geom density

Useful alternative to the histogram for continuous data that comes from an underlying smooth distribution.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()+
  ggtitle("Distribution of students by MATHS score")

ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()+
  ggtitle("Distribution of students by MATHS score and by gender")
```

### Geom boxplot

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()+
  ggtitle("Distribution of MATHS score for each gender")
```

### Notched plot

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)+
  ggtitle("Distribution of MATHS score for each gender")
```

Observation: Median MATHS score for female and male is different.

### Geom violin

Used to compare multiple data distributions

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()+
  ggtitle("Distribution of MATHS score by gender")
```

### Geom point aka scatterplot

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()+
  ggtitle("Distribution of students according to MATHS and ENGLISH scores")
```

### Combination of geom objects

Combination of geom boxplot and geom point (scatterplot)

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)+
  ggtitle("Distribution of MATHS scores by gender")
```

### Exploring statistical functions

Adding a statistical summary use stat_summary() method (overrides the default geom)

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun.y="mean",         
               colour ="light blue",        
               size=4)+
  ggtitle("Distribution of MATHS scores by gender with stat summary")
```

Adding a statistical summary using geom() method (overrides default stat)

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun.y="mean",           
             colour ="pink",          
             size=4)+
  ggtitle("Distribution of MATHS scores by gender with stat summary")
```

### Adding a best fit curve to scatterplot

Default loess smoothing method

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(linewidth=0.5)+
  ggtitle("Distribution of students by ENGLISH and MATHS scores")
```

Using lm smoothing method

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)+
  ggtitle("Distribution of students by ENGLISH and MATHS scores")
```

### Working with facets

Facetting is helpful in displaying different subsets of the data.

-   Using facet_wrap()

Better use of screen space than facet_grid because most displays are roughly rectangular. 
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="orange") +
    facet_wrap(~ CLASS)+
  ggtitle("Distribution of MATHS scores by class")
```
-   Using facet_grid()

Most useful when you have two discrete variables, and all combinations of the variables exist in the data.
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="purple") +
    facet_grid(~ CLASS)+
  ggtitle("Distribution of MATHS scores by class")
```
Main difference between the two functions:

- The facet_grid() function will produce a grid of plots for each combination of variables that you specify, even if some plots are empty.
- The facet_wrap() function will only produce plots for the combinations of variables that have values, which means it won’t produce any empty plots.

Source: https://www.statology.org/facet_wrap-vs-facet_grid-in-r/

### Coordinates
- Flip bar chart configuration from vertical to horizontal
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()+
  ggtitle("Distribution of students by race")

ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()+
  coord_flip()+
  ggtitle("Distribution of students by race")
```

### Changing x- and y- axis range
Chart below is misleading as x- and y-axis ranges are not equal. y-axis is denoted in chunks of 20 while x-axis is denoted in chunks of 25.
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, linewidth=0.5)+
  ggtitle("Distribution of students by ENGLISH and MATHS scores")
```

Adjusting the x- and y-axis to range from 0 to 100 in equal chunks
```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))+
  ggtitle("Distribution of students by ENGLISH and MATHS scores")
```

### Exploring themes in ggplot2
Themes control elements of the graph not related to the data i.e. background color, size of fonts and color of labels

- Theme grey
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()+
  ggtitle("Distribution of students by race")
```
- Theme classic
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()+
  ggtitle("Distribution of students by race")
```
- Theme minimal
```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()+
  ggtitle("Distribution of students by race")
```